Fixing Vulnerabilities Hinders Maintainability

### 2020-01-31 paper/icpc20.tex
new version

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
 static SUB_STATE_RETURN read_state_machine(SSL *s){ /*!\annotation{lst:func1}!*/
 // [snip]
+   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
+        && !BUF_MEM_grow_clean(s->init_buf, 
+                                (int)s->s3->tmp.message_size
+                                + SSL3_HM_HEADER_LENGTH)) { /*!\annotation{lst:func2}!*/
+         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
+         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+          return SUB_STATE_ERROR;
 // [snip]
 }
\end{lstlisting}

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2016-6307 vulnerability for file ssl/statem/statem\_lib.c},label={lst:vuln2}]
 int tls_get_message_header(SSL *s, int *mt){  /*!\annotation{lst:func3}!*/
 // [snip]
  l = RECORD_LAYER_get_rrec_length(&s->rlayer) + 
        SSL3_HM_HEADER_LENGTH;
- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func4}!*/
-      SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-      goto err;
   }
 // [snip]

@@

Although checking the maximum size of a message seems an elementary
problem to repair, the example shows that there was a significant amount of
changes performed in the code base which produced a negative impact in the
maintainability. The developer tries to patch the
code base by introducing a new branch point to the function \texttt{read\_state\_machine}~\ref{lst:func1} which adds complexity to the software
and disrupts
one of the guidelines proposed by the Software Improvement Group (SIG) for
building maintainable software~\cite{Visser:2016:OREILLY} $-$
the \emph{Write Simple Units of Code} guideline $-$ which resulted in a negative impact on the
OpenSSL maintainability.
%%%%%%% 
Software maintainability is designed as the degree to which an application is understood, repaired or enhanced. In this paper, our concern is to study whether, while improving software
security, developers are also negatively impacting the maintainability of their
software applications. This is important because software maintainability is approximately $75\%$ of the cost related to a project. To answer the following three research questions, we use two datasets
of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} to measure the impact of
security patches on the maintainability of open-source software. 
\begin{framed}
\textit{\textbf{RQ1} What is the impact of security patches on the
maintainability of open-source software?}
\end{framed}
\vspace{-0.1cm}
Often security flaws require patching code to make software more secure.
However, \textbf{there is no evidence yet of how security patches impact the
maintainability of open-source software}. Our suspicion is that developers tend
to introduce technical debt in their software when patching the code base to
address a security flaw because they tend to choose the easiest path to solve
it. To address it, we compute the maintainability of $1037$ commits using the
\emph{Better Code Hub} tool. 
\begin{framed}
\textit{\textbf{RQ2} Which patterns of security patches are more likely to
affect open-source software maintainability?}
\end{framed}
\vspace{-0.1cm}

@@

@@

\begin{framed}
\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
maintainability of open-source software?}
\end{framed}
\vspace{-0.1cm}

@@

@@

 	 	     	           \textbf{MISC} & 
Miscellaneous (673 cases) & 
This pattern comprises
 		several other security refactorings that do not have a CWE assigned and patterns that
 		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.\\
\bottomrule
\end{tabular}
\label{tab:patterns}
\end{table}

@@

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}
 	\caption{Maintainability report of OpenSSL's CVE-$2014$-$3506$ vulnerability
refactoring for the guideline \emph{Write Simple Units of Code} provided by
\emph{Better Code Hub}. This version of OpenSSL does not comply with the
guideline in the example since the bars do not reach the threshold points. This
example only complies with $\sfrac{1}{10}$ guidelines (\emph{Write Clean Code}).}
	\label{fig:bchrep}
\end{figure}

@@

\textbf{Write Clean Code} & Avoid producing software with code smells because
it is more likely to be maintainable in the future & \textbf{Code Smells:} 
\# of Occurrences~\cite{Visser:2016:OREILLY} (e.g., magic constants and long 
identifier names) \\
\bottomrule
\end{tabular}
\label{tab:guidelines}
\end{table}

@@

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.75\linewidth]{figures/commit.pdf}
 	\caption{Maintainability difference for security commits}
	\label{fig:commit}
\end{figure}

@@

\begin{framed}
\textit{\textbf{RQ1} What is the impact of security patches on the
maintainability of open-source software?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.1cm}

\begin{figure}[h]
 	\centering
 	\includegraphics[width=0.5\textwidth]{figures/impact_per_guideline.pdf}
 	\caption{Impact of the security patches for each guideline.}
	\label{fig:guidelines}
	\scalebox{0.5}{
	\begin{tabular}{r@{: }l r@{: }l}
	$WCC$ & Write Clean Code & $KACB$ & Keep Architecture Components Balanced\\
	$CACL$ & Couple Architecture Components Loosely & $SCM$ & Separate Concerns in Modules \\
	$KUIS$ & Keep Unit Interfaces Small & $WCO$ & Write Code Once \\
	$WSimpleUC$ & Write Simple Units of Code & $WShortUC$ & Write Short Units of Code
	\end{tabular}
	}
\end{figure}

@@

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.45\textwidth]{figures/maintainability_severity.pdf}
 	\caption{Maintainability difference by the vulnerabilities severity.}
	\label{fig:severity}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/maintainability_language.pdf}
  \caption{Maintainability difference by language after security patches.}
  \label{fig:lang_main}  
\end{figure}

@@

\begin{framed}
\textit{\textbf{RQ2} Which patterns of security patches are more likely to
affect open-source software maintainability?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.1cm}

@@

\begin{figure}[h]
 	\centering
 	\includegraphics[width=0.45\textwidth]{figures/maintainability_cwe.pdf}
 	\caption{Maintainability difference by pattern after security patches.}
	\label{fig:pat}	
\end{figure}

\begin{framed}
\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
maintainability of open-source software?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.1cm}

@@

\begin{figure}[h]	
 	\centering 	\includegraphics[width=0.43\textwidth]{figures/maintainability_sec_vs_reg.pdf}
 	\caption{Maintainability difference for security and baseline patches.}
	\label{fig:secvsreg}
\end{figure}

@@

\begin{framed}
\textit{\textbf{RQ1} What is the impact of security patches on the maintainability
of open-source software?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.1cm}

\textbf{\textit{Security patches harm the maintainability of open-source software.}}

@@

\begin{framed}
\textit{\textbf{RQ2} Which patterns of security patches are more likely to
affect open-source software maintainability?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.1cm}

\textbf{\textit{Path Traversal (CWE-22), Information Exposure (CWE-200), Resource Management Errors (CWE-399) are more likely to affect
open-source software maintainability.}}

@@

\begin{framed}
\textit{\textbf{RQ3} What is the impact of security patches versus regular changes
on the maintainability of open-source software?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.1cm}
\textbf{\textit{Patching security vulnerabilities has a larger negative and positive impact
than when fixing regular commits.}}

>>>>

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2016-6307 vulnerability for file ssl/statem/statem.c},label={lst:vuln}]
 static SUB_STATE_RETURN read_state_machine(SSL *s){ /*!\annotation{lst:func1}!*/
 // [snip]
+   if (!SSL_IS_DTLS(s) && s->s3->tmp.message_size > 0
+        && !BUF_MEM_grow_clean(s->init_buf, 
+                                (int)s->s3->tmp.message_size
+                                + SSL3_HM_HEADER_LENGTH)) { /*!\annotation{lst:func2}!*/
+         ssl3_send_alert(s, SSL3_AL_FATAL, SSL_AD_INTERNAL_ERROR);
+         SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
+          return SUB_STATE_ERROR;
 // [snip]
 }
\end{lstlisting}
 \vspace{-0.5cm}
\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={CStyle}, caption={Fix provided by OpenSSL developers to the
\\CVE-2016-6307 vulnerability for file ssl/statem/statem\_lib.c},label={lst:vuln2}]
 int tls_get_message_header(SSL *s, int *mt){  /*!\annotation{lst:func3}!*/
 // [snip]
  l = RECORD_LAYER_get_rrec_length(&s->rlayer) + 
        SSL3_HM_HEADER_LENGTH;
- if (l && !BUF_MEM_grow_clean(s->init_buf, (int)l)) { /*!\annotation{lst:func4}!*/
-      SSLerr(SSL_F_TLS_GET_MESSAGE_HEADER, ERR_R_BUF_LIB);
-      goto err;
   }
 // [snip]

@@

%
Although checking the maximum size of a message seems an elementary
problem to repair, the example shows that there was a significant amount of
changes performed in the code base which produced a negative impact in the
maintainability. The developer tries to patch the
code base by introducing a new branch point to the function \texttt{read\_state\_machine}~\ref{lst:func1} which adds complexity to the software
and disrupts
one of the guidelines proposed by the Software Improvement Group (SIG) for
building maintainable software~\cite{Visser:2016:OREILLY} $-$
the \emph{Write Simple Units of Code} guideline $-$ which resulted in a negative impact on the
OpenSSL maintainability.
%%%%%%% 
Software maintainability is designed as the degree to which an application is understood, repaired or enhanced. In this paper, our concern is to study whether, while improving software
security, developers are also negatively impacting the maintainability of their
software applications. This is important because software maintainability is approximately $75\%$ of the cost related to a project. To answer the following three research questions, we use two datasets
of security patches~\cite{Reis:2017:IJSSE, 10.1109/MSR.2019.00064} to measure the impact of
security patches on the maintainability of open-source software. 
%

\subsubsection*{\textit{\textbf{RQ1} What is the impact of security patches on the
maintainability of open-source software?}}

Often security flaws require patching code to make software more secure.
However, \textbf{there is no evidence yet of how security patches impact the
maintainability of open-source software}. Our suspicion is that developers tend
to introduce technical debt in their software when patching the code base to
address a security flaw because they tend to choose the easiest path to solve
it. To address it, we compute the maintainability of $1037$ commits using the
\emph{Better Code Hub} tool. 
%
\subsubsection*{\textit{\textbf{RQ2}  Which patterns of security patches are more likely to
affect open-source software maintainability?}}

@@

\vspace{-0.3cm}

@@

\subsubsection*{\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
maintainability of open-source software?}}

@@

	\vspace{-0.5cm}

@@

 	 	     	           \textbf{MISC} & 
Miscellaneous (673 cases) & 
This pattern comprises
 		several other security refactorings that do not have a CWE assigned and patterns that
 		do not satisfy the Wilcoxon test size requirement of more than $20$ refactorings.\\
\bottomrule
\end{tabular}
\label{tab:patterns}
\vspace{-0.5cm}
\end{table}

@@

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.45\textwidth]{figures/bch_report.png}
 	\caption{Maintainability report of OpenSSL's CVE-$2014$-$3506$ vulnerability
refactoring for the guideline \emph{Write Simple Units of Code} provided by
\emph{Better Code Hub}. This version of OpenSSL does not comply with the
guideline in the example since the bars do not reach the threshold points. This
example only complies with $\sfrac{1}{10}$ guidelines (\emph{Write Clean Code}).}
	\label{fig:bchrep}
	\vspace{-0.8cm}
\end{figure}

@@

\textbf{Write Clean Code} & Avoid producing software with code smells because
it is more likely to be maintainable in the future & \textbf{Code Smells:} 
\# of Occurrences~\cite{Visser:2016:OREILLY} (e.g., magic constants and long 
identifier names) \\
\bottomrule
\end{tabular}
\label{tab:guidelines}
\vspace{-0.3cm}
\end{table}

@@

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.75\linewidth]{figures/commit.pdf}
 	\caption{Maintainability difference for security commits}
	\label{fig:commit}
	\vspace{-0.5cm}
\end{figure}

@@

\subsubsection*{\textit{\textbf{RQ1} What is the impact of security patches on the
maintainability of open-source software?}}

\begin{figure}[h]
 	\centering
 	\includegraphics[width=0.5\textwidth]{figures/impact_per_guideline.pdf}
 	\caption{Impact of the security patches for each guideline.}
	\label{fig:guidelines}
	\scalebox{0.5}{
	\begin{tabular}{r@{: }l r@{: }l}
	$WCC$ & Write Clean Code & $KACB$ & Keep Architecture Components Balanced\\
	$CACL$ & Couple Architecture Components Loosely & $SCM$ & Separate Concerns in Modules \\
	$KUIS$ & Keep Unit Interfaces Small & $WCO$ & Write Code Once \\
	$WSimpleUC$ & Write Simple Units of Code & $WShortUC$ & Write Short Units of Code
	\end{tabular}
	}
	\vspace{-0.3cm}
\end{figure}

@@

\begin{figure}[h]
 	\centering 	\includegraphics[width=0.45\textwidth]{figures/maintainability_severity.pdf}
 	\caption{Maintainability difference by the vulnerabilities severity.}
	\label{fig:severity}
	\vspace{-0.4cm}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.5\textwidth]{figures/maintainability_language.pdf}
  \caption{Maintainability difference by language after security patches.}
  \label{fig:lang_main}  
  \vspace{-0.5cm}
\end{figure}

@@

\subsubsection*{\textit{\textbf{RQ2} Which patterns of security patches are more likely to
affect open-source software maintainability?}}

@@

\begin{figure}[h]
 	\centering
 	\includegraphics[width=0.45\textwidth]{figures/maintainability_cwe.pdf}
 	\caption{Maintainability difference by pattern after security patches.}
	\label{fig:pat}	
	\vspace{-0.5cm}
\end{figure}

\subsubsection*{\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
maintainability of open-source software?}}

@@

\begin{figure}[h]	
 	\centering 	\includegraphics[width=0.43\textwidth]{figures/maintainability_sec_vs_reg.pdf}
 	\caption{Maintainability difference for security and baseline patches.}
	\label{fig:secvsreg}
	\vspace{-0.5cm}
\end{figure}

@@

\subsubsection*{\textit{\textbf{RQ1} \textbf{What is the impact of security patches on the maintainability
of open-source software?}}}

\textit{Security patches harm the maintainability of open-source software.}

@@

\subsubsection*{\textit{\textbf{RQ2} \textbf{Which patterns of security patches are more likely to
affect open-source software maintainability?}}}

\textit{Path Traversal (CWE-22), Information Exposure (CWE-200), Resource Management Errors (CWE-399) are more likely to affect
open-source software maintainability.}

@@

\subsubsection*{\textit{\textbf{RQ3} \textbf{What is the impact of security patches versus regular changes
on the maintainability of open-source software?}}}
\textit{Patching security vulnerabilities has a larger negative and positive impact
than when fixing regular commits.}
